Distinguishing Spin-Aligned and Isotropic Black Hole Populations With Gravitational Waves

### 2017-05-10 paper/AlignedIsotropicGW.tex
Initial transition to nature formatting, some problems remain.

\documentclass[modern,linenumbers]{aastex61}

\usepackage{acronym}
\usepackage{amsmath}
\usepackage{amssymb}

@@

@@

\author[0000-0003-1540-8562]{Will M. Farr}

\affiliation{Birmingham Institute for Gravitational Wave Astronomy and
  School of Physics and Astronomy, University of Birmingham,
  Birmingham, B15 2TT, United Kingdom}

\author[0000-0002-6100-537X]{Simon Stevenson}

\affiliation{Birmingham Institute for Gravitational Wave Astronomy and
  School of Physics and Astronomy, University of Birmingham,
  Birmingham, B15 2TT, United Kingdom}
\affiliation{Kavli Institute for Theoretical Physics, Santa Barbara, CA 93106}

\author{M. Coleman Miller}

\affiliation{Department of Astronomy and Joint Space-Science
  Institute, University of Maryland, College Park, MD 20742--2421,
  United States}

\author[0000-0002-6134-8946]{Ilya Mandel}

\affiliation{Birmingham Institute for Gravitational Wave Astronomy and
  School of Physics and Astronomy, University of Birmingham,
  Birmingham, B15 2TT, United Kingdom}
\affiliation{Kavli Institute for Theoretical Physics, Santa Barbara, CA 93106}

\author[0000-0002-2916-9200]{Ben Farr}

\affiliation{Enrico Fermi Institute and Kavli Institute for Cosmological Physics, University of Chicago, Chicago, IL 60637, United States}

\author[0000-0002-6254-1617]{Alberto Vecchio}

\affiliation{Birmingham Institute for Gravitational Wave Astronomy and
  School of Physics and Astronomy, University of Birmingham,
  Birmingham, B15 2TT, United Kingdom}

\email{w.farr@bham.ac.uk, simons@star.sr.bham.ac.uk, miller@astro.umd.edu,\\
imandel@star.sr.bham.ac.uk, farr@uchicago.edu, av@star.sr.bham.ac.uk}

\begin{abstract}
  The first direct detections of \acp{GW}
  \citep{2016PhRvL.116f1102A,2016PhRvL.116x1103A,O1-BBH} from merging
  \acp{BBH} open a unique window into the \ac{BBH} formation
  environment.  One promising signature of the formation environment
  is the angular distribution of the \ac{BH} spins; systems formed
  through dynamical interactions among already-compact objects are
  expected to have isotropic spin orientations
  \citep{SigurdssonHernquist:1993,1993Natur.364..421K,PZMcMillan:2000,Rodriguez:2015,Stone:2016,2016ApJ...832L...2R}
  whereas binaries formed from pairs of stars born together are more
  likely to have spins preferentially aligned with the binary orbit as
  a consequence of their joint evolution toward a \ac{BBH} system
  \citep{TutukovYungelson:1973,TutukovYungelson:1993,Lipunov:1997,2016Natur.534..512B,Stevenson:2017,MandeldeMink:2016,Marchant:2016}.
  We consider existing \ac{GW} measurements of $\chieff$, the
  best-measured combination of spin parameters
  \citep{2016PhRvL.116x1102A,O1-BBH}, in the three likely binary black
  hole detections GW150914, LVT151012, and GW151226.  If binary black
  hole spin magnitudes extend to high values, as is suggested by
  observations of black hole X-ray binaries
  \citep{2015PhR...548....1M}, we show that the data already exhibit a
  $\OOneSigmaIsoAligned\sigma$ ($\OOneOddsIsoAligned$ odds
  ratio\footnote{An odds ratio of $r$ with $r \ll 1$ is equivalent to
    $x \sigma$ with $x = \Phi^{-1}\left( 1 - r/2 \right)$, where
    $\Phi$ is the unit normal CDF.})  preference for an isotropic
  angular distribution.  By considering the effect of 10 additional
  detections \citep{2016ApJ...833L...1A} drawn from the various models
  in the suite we show that if all observations come from a single
  population such an augmented data set would enable at least a
  $\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$ odds
  ratio) distinction between the isotropic and aligned models for the
  assumed spin magnitude distributions, and in most cases better than
  $5\sigma$ ($2.9 \times 10^{-7}$ odds ratio).  The existing
  preference for either an isotropic spin distribution or low spin
  magnitudes for the observed systems will be confirmed (or
  overturned) confidently in the near future by subsequent
  observations.
\end{abstract}

@@

Following the detection of a merging \ac{BBH} system, \ac{PE} tools
\citep{2015PhRvD..91d2003V} compare model gravitational waveforms
\citep[e.g.\
][]{2014PhRvD..89h4006P,2014PhRvD..89f1502T,2014PhRvL.113o1101H}
against the observed data to obtain a posterior distribution on the
parameters that describe the compact binary source.

The spin parameter with the largest effect on waveforms, and a
correspondingly tight constraint from the data, is a mass-weighted,
aligned, ``effective spin,''
\begin{equation}
  \chieff = \frac{c}{GM} \left( \frac{\vec{S}_1}{m_1} + \frac{\vec{S}_2}{m_2}
  \right) \cdot \frac{\vec{L}}{\left| \vec{L} \right|} \equiv \frac{1}{M} \left( m_1 \chi_1 + m_2 \chi_2 \right),
\end{equation}
where $m_{1,2}$ are the gravitational masses of the more-massive (1)
and less-massive (2) components, $M = m_1 + m_2$ is the total mass,
$\vec{S}_{1,2}$ are the spin angular momentum vectors of the black
holes in the binary, $\vec{L}$ is the orbital angular momentum vector,
assumed to point in the $\hat{z}$ direction, and
$0 \leq \chi_{1,2} \leq 1$ are the corresponding dimensionless
projections of the individual \ac{BH} spins
\citep{2016PhRvL.116x1102A}.

Figure \ref{fig:O1-posteriors} shows an approximation to the posterior
inferred on $\chieff$ for the three likely \ac{GW} detections
GW150914, GW151226 and LVT151012 from Advanced LIGO's first observing
run (O1) \citep{O1-BBH}.  Because samples drawn from the posterior on
$\chieff$ are not publicly released at this time, we have approximated
the posterior as a Gaussian distribution with the same mean and 90\%
credible interval as quoted in \citet{O1-BBH}.  None of the $\chieff$
posteriors are consistent with two black holes with large aligned
spins, $\chi_{1,2} \gtrsim 0.5$; this contrasts with the large spins
inferred for the majority of black holes in X-ray binaries with claimed spin measurements (see
Section \ref{sec:discussion}).  The analysis here is relatively
insensitive to the precise details of the posterior distributions;
other conclusions are more sensitive.  In particular, our Gaussian
approximation does permit $\chieff = 0$ for GW151226 while the true
posterior rules this out at high confidence
\citep{2016PhRvL.116x1103A,O1-BBH}.

\begin{figure}
  \plotone{../plots/chi-eff-mock-posteriors}
  \caption{\label{fig:O1-posteriors} Approximate posteriors on
    $\chieff$ from the Advanced LIGO O1 run observations in
    \citet{O1-BBH}.  We approximate the posteriors reported in
    \citet{O1-BBH} using Gaussians with the same median and 90\%
    credible interval as reported in \citet{O1-BBH}.  It is notable
    that none of the $\chieff$ posteriors support high \ac{BH}
    spin magnitudes with aligned spins, suggested by observations of stellar-mass black holes in  X-ray binaries (see
    \citet{2015PhR...548....1M} for a summary of such measurements).}
\end{figure}

@@

Each of our models for the merging \ac{BBH} spin population assumes
that the merging black holes are of equal mass (this is marginally
consistent with the three observations \citep{O1-BBH}, and the
$\chieff$ distribution is not particularly sensitive to the mass ratio
between the merging objects---see Section \ref{sec:mass-ratio}).  We
assume that the population spin distribution factorises into a
distribution for the spin magnitude $a$ and a distribution for the
spin angles.  Finally, we assume that the distribution of spins is
common to each component in a merging binary.  Choosing one of three
magnitude distributions
\begin{equation}
  \label{eq:magnitude-dists}
  p(a) = \begin{cases}
    2\left(1-a \right) & \textnormal{``low''} \\
    1 & \textnormal{``flat''} \\
    2 a & \textnormal{``high''}
  \end{cases},
\end{equation}
and pairing with either an isotropic angular distribution or a
distribution that generates perfect alignment with the positive $z$
axis yields six different models for the $\chieff$ distribution.
These models are shown in Figure \ref{fig:chieff-distribution-models}.

@@

The distributions in Eq.\ \eqref{eq:magnitude-dists} are not meant to
represent any particular physical model, but rather to capture our
uncertainty about the spin magnitude distribution, discussed in detail
in Section \ref{sec:discussion}; neither observations nor population
synthesis codes can at this point authoritatively suggest \emph{any}
particular spin distribution \citep{2015PhR...548....1M}.  Our models,
however, allow us to see how sensitive the $\chieff$ distribution is
to spin alignment given uncertainties about the spin magnitudes.

We fit hierarchical models of the three LIGO O1 observations using
these six different, zero-parameter population distributions (see
Section \ref{sec:hierarchical}).  We also fit three mixture models for
the population, where the spin magnitude distribution is fixed but the
angular distribution is a weighted sum of the isotropic and aligned
distribution.  The evidence, or marginal likelihood, for each of the
models is shown in Figure \ref{fig:O1-odds}.  For all three magnitude
distributions, the mixture models' posterior on the mixing fraction
peaks at 100\% isotropic, which explains why the zero-parameter,
pure-isotropic models are preferred over the single-parameter mixture
models for every magnitude distribution with this data set.  Not
surprisingly, given the small $\chieff$ values in the three detected
systems, the most-favoured model among those with an isotropic angular
distribution has the ``low'' magnitude distribution; the most favoured
model among those with an aligned distribution also has the ``low''
magnitude distribution.  The odds ratio between the ``low'' aligned
and ``low'' isotropic models is $\OOneOddsIsoAligned$, or
$\OOneSigmaIsoAligned\sigma$; thus the data favour isotropic spins
among our suite of models.  While the data favour spin amplitude
distributions with small spin magnitudes, note that a model with all
\ac{BBH} systems having zero spin is ruled out by the GW151226
measurements, which bound at least one black hole to have spin
magnitude $\geq 0.2$ at 99\% credibility \citep{2016PhRvL.116x1103A}.

@@

Estimates of the rate of \ac{BBH} coalescences give a reasonable
chance of 10 additional \ac{BBH} detections in the next two years
\citep{O1-BBH,2016ApJ...833L...1A,2016ApJS..227...14A}.  Assuming 10
additional detections drawn from each of our six zero-parameter models
for the spin distribution in addition to the three existing detections
from O1, with observational uncertainties drawn randomly from the
three Gaussian widths used to approximate the $\chieff$ posteriors in
Figure \ref{fig:O1-posteriors}\footnote{The measurement uncertainty in
  $\chieff$ depends on the other parameters of the merging \ac{BBH}
  system, particularly on the mass ratio.  Our assumption about future
  observational uncertainties is appropriate if the parameters of the
  three detected events are representative of the parameters of future
  detections. See Section \ref{sec:chi-eff-precision}
for further discussion.}, we find the odds ratios shown in Figure
\ref{fig:O2-predictions}.  We find that most scenarios with an
additional 10 detections allow the simulated angular distribution to
be inferred with greater than $5\sigma$ ($2.9 \times 10^{-7}$ odds)
credibility; and in the most pessimistic case the distinction is
typically $\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$
odds ratio).  While such future detections should permit a confident
distinction between \emph{angular} distributions, we would remain much
less certain about the \emph{magnitude} distribution among the three
options considered here until we have a larger number of observations.
In Figure \ref{fig:O2-predictions}, the odds ratio between different
magnitude distributions with the same angular distribution is much
closer to unity than the odds ratio between angular distributions.

@@

Most of our resolving power for the spin angular distribution is a
result of the fact that our ``aligned'' models cannot produce
$\chieff < 0$ (see Figure \ref{fig:chieff-distribution-models}).  If
spins are intrinsically very small, with $a \lesssim 0.3$, then it is
no longer possible to resolve the negative effective spin with a small
number of observations.  As noted below, however, spins observed in
X-ray binaries are typically large.  Additionally, models which do not
permit \emph{some} spins with $\chieff \gtrsim 0.1$ are ruled out by
the GW151226 observations \citep{2016PhRvL.116x1103A}.

In order to perform our analysis we need to select models for the
distribution of the spin magnitudes of stellar-mass black holes.
Observational data for such model selection is sparse.
\citet{2015PhR...548....1M} give current estimates of the spin
parameters for stellar-mass black holes, obtained using disk
reflection and/or disk continuum methods.  Most of the systems studied
are low-mass X-ray binaries rather than the high-mass X-ray binaries
that are likely to be the progenitors of double black hole binaries.
In addition, there are substantial systematic errors that can
complicate either type of analysis (see \citealt{2015PhR...548....1M}
for a discussion), and selection effects could yield a biased
distribution. Nonetheless, if we take the reported spin magnitudes as
representative then we find that there is a preference for high spins;
for example, 14 of the 19 systems with reported spins have
dimensionless spin parameters in excess of 0.5.  It is usually argued
that the masses and spin parameters of stellar-mass black holes are
unlikely to be altered significantly by accretion (low-mass donors may
not have enough mass and high-mass donors have a very short phase in
which they transfer mass; a variant of this long-standing argument was
presented by \citet{1999MNRAS.305..654K}), but see
\citet{2003MNRAS.341..385P,2015ApJ...800...17F}.  Thus the current
spin parameters probably are close to their values upon core collapse,
at least in high-mass X-ray binaries.  However, the specific processes
involved in the production of black hole binaries from isolated
binaries could alter the spin magnitude distribution of those holes
relative to the X-ray binary systems; for example, close tidal
interactions could spin up the core, or stripping of the envelope
could reduce the available angular momentum
\citep{2016MNRAS.462..844K,2017arXiv170200885Z,2017arXiv170203952H}.

The spin directions of binary black holes formed dynamically through
interactions in dense stellar environments
\citep{SigurdssonHernquist:1993,1993Natur.364..421K,PZMcMillan:2000,Rodriguez:2015,Stone:2016}
are expected to be isotropic given the absence of a preferred
direction \citep[e.g.,][]{2016ApJ...832L...2R} and the persistence of
an isotropic distribution through post-Newtonian evolution
\citep{2004PhRvD..70l4020S,2007ApJ...661L.147B}.

The spin directions in isolated binaries, whether evolving via the
classical channel through a common-envelope phase
\citep{TutukovYungelson:1973,TutukovYungelson:1993,Lipunov:1997,2016Natur.534..512B,Stevenson:2017}
or through chemically homogeneous evolution
\citep{MandeldeMink:2016,Marchant:2016} are usually expected to be
preferentially aligned.  Despite observed spin-orbit misalignments in
both massive stellar binaries
\citep{Albrecht:2009,2014ApJ...785...83A} and BH X-ray binaries
\citep{Orosz:2001,Martin:2008b,Martin:2008,MorningstarMiller:2014},
mass accretion and tidal interactions will tend to realign the binary.
On the other hand, a supernova natal kick (if any) can change the
orbital plane and misalign the binary
\citep{2000ApJ...541..319K,2013PhRvD..87j4028G}; the supernova can
also tilt the spin angle, as in the double pulsar J0737-3039
\citep{2011ApJ...742...81F}; and a variety of uncertain processes,
such as wind-driven mass loss and post-collapse fallback, can couple
the spin magnitude and direction distributions, contrary to our
simplified assumptions.  A small misalignment at wide separation can
also evolve to a more significant misalignment in component spins as
the binary spirals in through \ac{GW} emission \citep{2015PhRvD..92f4016G},
but $\chieff$ is approximately conserved through this evolution.

\citet{2017CQGra..34cLT01V} also studied the possibility of
distinguishing aligned and isotropic angular distributions of \ac{BBH}
spins, but concluded that several hundred sources would be required to
adequately separate models which included both parallel and
anti-parallel spins for an ``aligned'' population.  In contrast, and
in agreement with the present study, \citet{Stevenson:2017spin} found
that only $\sim 5$ observations of rapidly spinning black holes would
be necessary to distinguish isotropic and aligned spin distributions,
though tens of detections would be required for more nuanced admixture
models.  Meanwhile, recent studies by
\citet{2017arXiv170306869F} and \citet{2017arXiv170306223G} focused on inference
on spin magnitudes.

@@

\acknowledgements

Supplementary Information is linked to the online version of the paper
at \url{http://www.nature.com/nature}.

We thank Richard O'Shaughnessy, Christopher Berry, and Davide Gerosa
for discussions and comments on this work.  WF, SS, IM and AV were
supported in part by the STFC.  MCM acknowledges support of the
University of Birmingham Institute for Advanced Study Distinguished
Visiting Fellows program.  SS and IM acknowledge support from the
National Science Foundation under Grant No. NSF PHY11-25915.

All authors contributed to the work presented in this paper.

Reprints and permissions information is available at
\url{http://www.nature.com/reprints}.

The authors declare no competing financial interests.

Correspondence and requests for materials should be addressed to
\href{mailto:w.farr@bham.ac.uk}{w.farr@bham.ac.uk}.

\appendix

@@

LIGO measures $\chieff$ better than any other spin parameter, but
still with significant uncertainty, so we need to properly incorporate
measurement uncertainty in our analysis; thus our analysis must be
\emph{hierarchical} \citep{2010ApJ...725.2166H,2010PhRvD..81h4029M}.
In a hierarchical analysis, we assume that each event has a true, but
unknown, value of the effective spin, drawn from the population
distribution, which may have some parameters $\lambda$; then the
system is observed, represented by the likelihood function, which
results in a distribution for the true effective spin (and all other
parameters describing the system) consistent with the data.
Combining, the joint posterior on each system's $\chieff^i$ parameters
and the population parameters $\lambda$ implied by a set of
observations each with data $d^i$, is
\begin{equation}
  p\left( \left\{ \chieff^i \right\}, \lambda \mid \left\{ d^i \right\} \right) \propto \left[ \prod_{i=1}^{N_\mathrm{obs}} p\left(d^i \mid \chieff^i \right) p\left( \chieff^i \mid \lambda \right) \right] p\left(\lambda\right).
\end{equation}

@@

\bibliography{AlignedIsotropicGW}

>>>>

\documentclass{nature}

\usepackage{acronym}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{graphicx}

@@

\newcommand{\plotone}[1]{\includegraphics[width=\columnwidth]{#1}}

% Journal Abbreviations
\newcommand{\aap}{Astronomy and Astrophysics}
\newcommand{\apj}{Astrophysical Journal}
\newcommand{\apjl}{Astrophysical Journal Letters}
\newcommand{\apjs}{Astrophysical Journal Supplements}
\newcommand{\mnras}{Monthly Notices of the Royal Astronomical Society}
\newcommand{\nat}{Nature}
\newcommand{\physrep}{Physics Reports}
\newcommand{\prd}{Physical Review D}

@@

\author{Will M. Farr$^1$, Simon Stevenson$^1$, M. Coleman Miller$^2$,
  Ilya Mandel$^{1,3}$, Ben Farr$^4$ \& Alberto Vecchio$^1$}

\maketitle

\begin{affiliations}
\item Birmingham Institute for Gravitational Wave Astronomy and
  School of Physics and Astronomy, University of Birmingham,
  Birmingham, B15 2TT, United Kingdom
\item Department of Astronomy and Joint Space-Science
  Institute, University of Maryland, College Park, MD 20742--2421,
  United States
\item Kavli Institute for Theoretical Physics, Santa Barbara, CA 93106
\item Enrico Fermi Institute and Kavli Institute for Cosmological
  Physics, University of Chicago, Chicago, IL 60637, United States
\end{affiliations}

\begin{abstract}
  The first direct detections of \acp{GW}
  \cite{2016PhRvL.116f1102A,2016PhRvL.116x1103A,O1-BBH} from merging
  \acp{BBH} open a unique window into the \ac{BBH} formation
  environment.  One promising signature of the formation environment
  is the angular distribution of the \ac{BH} spins; systems formed
  through dynamical interactions among already-compact objects are
  expected to have isotropic spin orientations
  \cite{SigurdssonHernquist:1993,1993Natur.364..421K,PZMcMillan:2000,Rodriguez:2015,Stone:2016,2016ApJ...832L...2R}
  whereas binaries formed from pairs of stars born together are more
  likely to have spins preferentially aligned with the binary orbit as
  a consequence of their joint evolution toward a \ac{BBH} system
  \cite{TutukovYungelson:1973,TutukovYungelson:1993,Lipunov:1997,2016Natur.534..512B,Stevenson:2017,MandeldeMink:2016,Marchant:2016}.
  We consider existing \ac{GW} measurements of $\chieff$, the
  best-measured combination of spin parameters
  \cite{2016PhRvL.116x1102A,O1-BBH}, in the three likely binary black
  hole detections GW150914, LVT151012, and GW151226.  If binary black
  hole spin magnitudes extend to high values, as is suggested by
  observations of black hole X-ray binaries
  \cite{2015PhR...548....1M}, we show that the data already exhibit a
  $\OOneSigmaIsoAligned\sigma$ ($\OOneOddsIsoAligned$ odds
  ratio\footnote{An odds ratio of $r$ with $r \ll 1$ is equivalent to
    $x \sigma$ with $x = \Phi^{-1}\left( 1 - r/2 \right)$, where
    $\Phi$ is the unit normal CDF.})  preference for an isotropic
  angular distribution.  By considering the effect of 10 additional
  detections \cite{2016ApJ...833L...1A} drawn from the various models
  in the suite we show that if all observations come from a single
  population such an augmented data set would enable at least a
  $\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$ odds
  ratio) distinction between the isotropic and aligned models for the
  assumed spin magnitude distributions, and in most cases better than
  $5\sigma$ ($2.9 \times 10^{-7}$ odds ratio).  The existing
  preference for either an isotropic spin distribution or low spin
  magnitudes for the observed systems will be confirmed (or
  overturned) confidently in the near future by subsequent
  observations.
\end{abstract}

@@

Following the detection of a merging \ac{BBH} system, \ac{PE} tools
\cite{2015PhRvD..91d2003V} compare model gravitational waveforms
\cite{2014PhRvD..89h4006P,2014PhRvD..89f1502T,2014PhRvL.113o1101H}
against the observed data to obtain a posterior distribution on the
parameters that describe the compact binary source.

The spin parameter with the largest effect on waveforms, and a
correspondingly tight constraint from the data, is a mass-weighted,
aligned, ``effective spin,''
\begin{equation}
  \chieff = \frac{c}{GM} \left( \frac{\vec{S}_1}{m_1} + \frac{\vec{S}_2}{m_2}
  \right) \cdot \frac{\vec{L}}{\left| \vec{L} \right|} \equiv \frac{1}{M} \left( m_1 \chi_1 + m_2 \chi_2 \right),
\end{equation}
where $m_{1,2}$ are the gravitational masses of the more-massive (1)
and less-massive (2) components, $M = m_1 + m_2$ is the total mass,
$\vec{S}_{1,2}$ are the spin angular momentum vectors of the black
holes in the binary, $\vec{L}$ is the orbital angular momentum vector,
assumed to point in the $\hat{z}$ direction, and
$0 \leq \chi_{1,2} \leq 1$ are the corresponding dimensionless
projections of the individual \ac{BH} spins
\cite{2016PhRvL.116x1102A}.

Figure \ref{fig:O1-posteriors} shows an approximation to the posterior
inferred on $\chieff$ for the three likely \ac{GW} detections
GW150914, GW151226 and LVT151012 from Advanced LIGO's first observing
run (O1) \cite{O1-BBH}.  Because samples drawn from the posterior on
$\chieff$ are not publicly released at this time, we have approximated
the posterior as a Gaussian distribution with the same mean and 90\%
credible interval as quoted in \cite{O1-BBH}.  None of the $\chieff$
posteriors are consistent with two black holes with large aligned
spins, $\chi_{1,2} \gtrsim 0.5$; this contrasts with the large spins
inferred for the majority of black holes in X-ray binaries with claimed spin measurements (see
Section \ref{sec:discussion}).  The analysis here is relatively
insensitive to the precise details of the posterior distributions;
other conclusions are more sensitive.  In particular, our Gaussian
approximation does permit $\chieff = 0$ for GW151226 while the true
posterior rules this out at high confidence
\cite{2016PhRvL.116x1103A,O1-BBH}.

\begin{figure}
  \plotone{../plots/chi-eff-mock-posteriors}
  \caption{\label{fig:O1-posteriors} Approximate posteriors on
    $\chieff$ from the Advanced LIGO O1 run observations in
    \cite{O1-BBH}.  We approximate the posteriors reported in
    \cite{O1-BBH} using Gaussians with the same median and 90\%
    credible interval as reported in \cite{O1-BBH}.  It is notable
    that none of the $\chieff$ posteriors support high \ac{BH}
    spin magnitudes with aligned spins, suggested by observations of stellar-mass black holes in  X-ray binaries (see
    \cite{2015PhR...548....1M} for a summary of such measurements).}
\end{figure}

@@

Each of our models for the merging \ac{BBH} spin population assumes
that the merging black holes are of equal mass (this is marginally
consistent with the three observations \cite{O1-BBH}, and the
$\chieff$ distribution is not particularly sensitive to the mass ratio
between the merging objects---see Section \ref{sec:mass-ratio}).  We
assume that the population spin distribution factorises into a
distribution for the spin magnitude $a$ and a distribution for the
spin angles.  Finally, we assume that the distribution of spins is
common to each component in a merging binary.  Choosing one of three
magnitude distributions
\begin{equation}
  \label{eq:magnitude-dists}
  p(a) = \begin{cases}
    2\left(1-a \right) & \textnormal{``low''} \\
    1 & \textnormal{``flat''} \\
    2 a & \textnormal{``high''}
  \end{cases},
\end{equation}
and pairing with either an isotropic angular distribution or a
distribution that generates perfect alignment with the positive $z$
axis yields six different models for the $\chieff$ distribution.
These models are shown in Figure \ref{fig:chieff-distribution-models}.

@@

The distributions in Eq.\ \eqref{eq:magnitude-dists} are not meant to
represent any particular physical model, but rather to capture our
uncertainty about the spin magnitude distribution, discussed in detail
in Section \ref{sec:discussion}; neither observations nor population
synthesis codes can at this point authoritatively suggest \emph{any}
particular spin distribution \cite{2015PhR...548....1M}.  Our models,
however, allow us to see how sensitive the $\chieff$ distribution is
to spin alignment given uncertainties about the spin magnitudes.

We fit hierarchical models of the three LIGO O1 observations using
these six different, zero-parameter population distributions (see
Section \ref{sec:hierarchical}).  We also fit three mixture models for
the population, where the spin magnitude distribution is fixed but the
angular distribution is a weighted sum of the isotropic and aligned
distribution.  The evidence, or marginal likelihood, for each of the
models is shown in Figure \ref{fig:O1-odds}.  For all three magnitude
distributions, the mixture models' posterior on the mixing fraction
peaks at 100\% isotropic, which explains why the zero-parameter,
pure-isotropic models are preferred over the single-parameter mixture
models for every magnitude distribution with this data set.  Not
surprisingly, given the small $\chieff$ values in the three detected
systems, the most-favoured model among those with an isotropic angular
distribution has the ``low'' magnitude distribution; the most favoured
model among those with an aligned distribution also has the ``low''
magnitude distribution.  The odds ratio between the ``low'' aligned
and ``low'' isotropic models is $\OOneOddsIsoAligned$, or
$\OOneSigmaIsoAligned\sigma$; thus the data favour isotropic spins
among our suite of models.  While the data favour spin amplitude
distributions with small spin magnitudes, note that a model with all
\ac{BBH} systems having zero spin is ruled out by the GW151226
measurements, which bound at least one black hole to have spin
magnitude $\geq 0.2$ at 99\% credibility \cite{2016PhRvL.116x1103A}.

@@

Estimates of the rate of \ac{BBH} coalescences give a reasonable
chance of 10 additional \ac{BBH} detections in the next two years
\cite{O1-BBH,2016ApJ...833L...1A,2016ApJS..227...14A}.  Assuming 10
additional detections drawn from each of our six zero-parameter models
for the spin distribution in addition to the three existing detections
from O1, with observational uncertainties drawn randomly from the
three Gaussian widths used to approximate the $\chieff$ posteriors in
Figure \ref{fig:O1-posteriors}\footnote{The measurement uncertainty in
  $\chieff$ depends on the other parameters of the merging \ac{BBH}
  system, particularly on the mass ratio.  Our assumption about future
  observational uncertainties is appropriate if the parameters of the
  three detected events are representative of the parameters of future
  detections. See Section \ref{sec:chi-eff-precision}
for further discussion.}, we find the odds ratios shown in Figure
\ref{fig:O2-predictions}.  We find that most scenarios with an
additional 10 detections allow the simulated angular distribution to
be inferred with greater than $5\sigma$ ($2.9 \times 10^{-7}$ odds)
credibility; and in the most pessimistic case the distinction is
typically $\OTwoSigmaIsoAlignedMin\sigma$ ($\OTwoOddsIsoAlignedMin$
odds ratio).  While such future detections should permit a confident
distinction between \emph{angular} distributions, we would remain much
less certain about the \emph{magnitude} distribution among the three
options considered here until we have a larger number of observations.
In Figure \ref{fig:O2-predictions}, the odds ratio between different
magnitude distributions with the same angular distribution is much
closer to unity than the odds ratio between angular distributions.

@@

Most of our resolving power for the spin angular distribution is a
result of the fact that our ``aligned'' models cannot produce
$\chieff < 0$ (see Figure \ref{fig:chieff-distribution-models}).  If
spins are intrinsically very small, with $a \lesssim 0.3$, then it is
no longer possible to resolve the negative effective spin with a small
number of observations.  As noted below, however, spins observed in
X-ray binaries are typically large.  Additionally, models which do not
permit \emph{some} spins with $\chieff \gtrsim 0.1$ are ruled out by
the GW151226 observations \cite{2016PhRvL.116x1103A}.

In order to perform our analysis we need to select models for the
distribution of the spin magnitudes of stellar-mass black holes.
Observational data for such model selection is sparse.
\cite{2015PhR...548....1M} give current estimates of the spin
parameters for stellar-mass black holes, obtained using disk
reflection and/or disk continuum methods.  Most of the systems studied
are low-mass X-ray binaries rather than the high-mass X-ray binaries
that are likely to be the progenitors of double black hole binaries.
In addition, there are substantial systematic errors that can
complicate either type of analysis (see \cite{2015PhR...548....1M} for
a discussion), and selection effects could yield a biased
distribution. Nonetheless, if we take the reported spin magnitudes as
representative then we find that there is a preference for high spins;
for example, 14 of the 19 systems with reported spins have
dimensionless spin parameters in excess of 0.5.  It is usually argued
that the masses and spin parameters of stellar-mass black holes are
unlikely to be altered significantly by accretion (low-mass donors may
not have enough mass and high-mass donors have a very short phase in
which they transfer mass; a variant of this long-standing argument was
presented by \cite{1999MNRAS.305..654K}), but see
\cite{2003MNRAS.341..385P,2015ApJ...800...17F}.  Thus the current spin
parameters probably are close to their values upon core collapse, at
least in high-mass X-ray binaries.  However, the specific processes
involved in the production of black hole binaries from isolated
binaries could alter the spin magnitude distribution of those holes
relative to the X-ray binary systems; for example, close tidal
interactions could spin up the core, or stripping of the envelope
could reduce the available angular momentum
\cite{2016MNRAS.462..844K,2017arXiv170200885Z,2017arXiv170203952H}.

The spin directions of binary black holes formed dynamically through
interactions in dense stellar environments
\cite{SigurdssonHernquist:1993,1993Natur.364..421K,PZMcMillan:2000,Rodriguez:2015,Stone:2016}
are expected to be isotropic given the absence of a preferred
direction \cite{2016ApJ...832L...2R} and the persistence of an
isotropic distribution through post-Newtonian evolution
\cite{2004PhRvD..70l4020S,2007ApJ...661L.147B}.

The spin directions in isolated binaries, whether evolving via the
classical channel through a common-envelope phase
\cite{TutukovYungelson:1973,TutukovYungelson:1993,Lipunov:1997,2016Natur.534..512B,Stevenson:2017}
or through chemically homogeneous evolution
\cite{MandeldeMink:2016,Marchant:2016} are usually expected to be
preferentially aligned.  Despite observed spin-orbit misalignments in
both massive stellar binaries
\cite{Albrecht:2009,2014ApJ...785...83A} and BH X-ray binaries
\cite{Orosz:2001,Martin:2008b,Martin:2008,MorningstarMiller:2014},
mass accretion and tidal interactions will tend to realign the binary.
On the other hand, a supernova natal kick (if any) can change the
orbital plane and misalign the binary
\cite{2000ApJ...541..319K,2013PhRvD..87j4028G}; the supernova can
also tilt the spin angle, as in the double pulsar J0737-3039
\cite{2011ApJ...742...81F}; and a variety of uncertain processes,
such as wind-driven mass loss and post-collapse fallback, can couple
the spin magnitude and direction distributions, contrary to our
simplified assumptions.  A small misalignment at wide separation can
also evolve to a more significant misalignment in component spins as
the binary spirals in through \ac{GW} emission \cite{2015PhRvD..92f4016G},
but $\chieff$ is approximately conserved through this evolution.

\cite{2017CQGra..34cLT01V} also studied the possibility of
distinguishing aligned and isotropic angular distributions of \ac{BBH}
spins, but concluded that several hundred sources would be required to
adequately separate models which included both parallel and
anti-parallel spins for an ``aligned'' population.  In contrast, and
in agreement with the present study, \cite{Stevenson:2017spin} found
that only $\sim 5$ observations of rapidly spinning black holes would
be necessary to distinguish isotropic and aligned spin distributions,
though tens of detections would be required for more nuanced admixture
models.  Meanwhile, recent studies by
\cite{2017arXiv170306869F} and \cite{2017arXiv170306223G} focused on inference
on spin magnitudes.

@@

\begin{addendum}
\item We thank Richard O'Shaughnessy, Christopher Berry, and Davide
  Gerosa for discussions and comments on this work.  WF, SS, IM and AV
  were supported in part by the STFC.  MCM acknowledges support of the
  University of Birmingham Institute for Advanced Study Distinguished
  Visiting Fellows program.  SS and IM acknowledge support from the
  National Science Foundation under Grant No. NSF PHY11-25915.
\item[Supplementary Information] Supplementary Information is linked
  to the online version of the paper at
  \url{http://www.nature.com/nature}.
\item[Author Contributions] All authors contributed to the work
  presented in this paper.
\item[Author Information] Reprints and permissions information is
  available at \url{http://www.nature.com/reprints}.  The authors
  declare no competing financial interests.  Correspondence and
  requests for materials should be addressed to
  \href{mailto:w.farr@bham.ac.uk}{w.farr@bham.ac.uk}.
\end{addendum}

\begin{methods}

@@

LIGO measures $\chieff$ better than any other spin parameter, but
still with significant uncertainty, so we need to properly incorporate
measurement uncertainty in our analysis; thus our analysis must be
\emph{hierarchical} \cite{2010ApJ...725.2166H,2010PhRvD..81h4029M}.
In a hierarchical analysis, we assume that each event has a true, but
unknown, value of the effective spin, drawn from the population
distribution, which may have some parameters $\lambda$; then the
system is observed, represented by the likelihood function, which
results in a distribution for the true effective spin (and all other
parameters describing the system) consistent with the data.
Combining, the joint posterior on each system's $\chieff^i$ parameters
and the population parameters $\lambda$ implied by a set of
observations each with data $d^i$, is
\begin{equation}
  p\left( \left\{ \chieff^i \right\}, \lambda \mid \left\{ d^i \right\} \right) \propto \left[ \prod_{i=1}^{N_\mathrm{obs}} p\left(d^i \mid \chieff^i \right) p\left( \chieff^i \mid \lambda \right) \right] p\left(\lambda\right).
\end{equation}

@@

\end{methods}

\bibliographystyle{naturemag}
\bibliography{AlignedIsotropicGW}
